Model Arrival Variation
========================================================

This document describes the model of Arrival/catch observed in Arizona and how it covaries with energetic demand and resource availability in Mexican wintering grounds

```{r}
#load the Hummingbird arrival data; i.e. the d.f. yearly.catch.stat
load("/home/sarah/Documents/GitHub/extreme_limits/data/yearly.catch.stat.rdata")
```

```{r}
#Calculate quantiles of arrival distribution per site & year ('sess.quant...'). quant.50 indicates the median arrival date at a given site and year.
#note that in the sess.quant variables the year is irrelevant
require(ggplot2)

yearly.catch.stat$sess.quant.05 <- as.Date(yearly.catch.stat$sess.quant.05, origin="2015-01-01")
yearly.catch.stat$sess.quant.25 <- as.Date(yearly.catch.stat$sess.quant.25, origin="2015-01-01")
yearly.catch.stat$sess.quant.50 <- as.Date(yearly.catch.stat$sess.quant.50, origin="2015-01-01")
yearly.catch.stat$sess.quant.75 <- as.Date(yearly.catch.stat$sess.quant.75, origin="2015-01-01")
yearly.catch.stat$sess.quant.95 <- as.Date(yearly.catch.stat$sess.quant.95, origin="2015-01-01")

ggplot(yearly.catch.stat, aes(yr, sess.quant.50)) + geom_point(aes(size=n.birds)) + theme_bw() + 
  stat_smooth() + scale_x_continuous(breaks=c(2001:2012))

range(yearly.catch.stat$sess.quant.50)
mean(yearly.catch.stat$sess.quant.50)
mean(yearly.catch.stat[yearly.catch.stat$yr==2006,]$sess.quant.50)
mean(yearly.catch.stat[yearly.catch.stat$yr==2010,]$sess.quant.50)
````

Load the yearly climate (and NDVI) data which will serve as predictors and merge it with the arrival/catch metrics in #the data frame 'yearly.catch.stat' to be used for modelling
```{r}
load("/home/sarah/Documents/GitHub/extreme_limits/data/yearly.climate.rdata")
#gives you yearly.climate
yearly.catch.stat <- merge(yearly.catch.stat, yearly.climate)
```

Plot relationship of NDVI with precipitation and with physiological demand
```{r}
require(ggplot2)

#precipitation correlated with NDVI in winter?
ggplot(yearly.climate, aes(wintercumpre, winterNDVI)) + geom_point() + stat_smooth(method="lm") + 
  ylab("average winter NDVI") + xlab("cumulative winter precip (mm)") + #geom_text(aes(label=yr)) +
  theme_classic() + theme(text = element_text(size=20))
lm1 = lm(winterNDVI ~ wintercumpre, data = yearly.climate)
print(summary(lm1)$r.sq)

#NDVI correlated with physiological demand?
ggplot(yearly.climate, aes(yearly.Te.10C.q, winterNDVI)) + geom_point() + stat_smooth(method="lm") + 
  ylab("average winter NDVI") + xlab("percent days with Te < 10 C") + geom_text(aes(label=yr)) +
  theme_classic() + theme(text = element_text(size=20))
lm2 = lm(winterNDVI ~ yearly.Te.10C.q, data = yearly.climate)
print(summary(lm2)$r.sq)

rm(yearly.climate)
```

Plot arrival responses to a proxy for physiological (yearly.Te.10C.q) demand and resrouce availability (NDVI)
```{r fig.width=7, fig.height=6, warning=FALSE}
require(lattice)
require(latticeExtra)
xyplot(sess.quant.50 ~ yearly.Te.10C.q|LOC, data=yearly.catch.stat) + layer(panel.smoother(x, y, method = "lm"))
xyplot(sess.quant.50 ~ winterNDVI|LOC, data=yearly.catch.stat) + layer(panel.smoother(x, y, method = "lm"))

xyplot(sess.quant.50 ~ yearly.Te.10C.q, data=yearly.catch.stat) + layer(panel.smoother(x, y, method = "lm"))
xyplot(sess.quant.50 ~ winterNDVI, data=yearly.catch.stat) + layer(panel.smoother(x, y, method = "lm"))

```


GGPLOT VERSION: Plot molt responses to a proxy for physiological (yearly.Te.10C.q) demand and resrouce availability (NDVI)
```{r fig.width=7, fig.height=6, warning=FALSE}
require(ggplot2)

ycs_cc = yearly.catch.stat[complete.cases(yearly.catch.stat),]

ggplot(ycs_cc, aes(yearly.Te.10C.q, sess.quant.50)) + geom_point() + theme_bw() + 
  stat_smooth(method="lm", formula = y ~ x, aes(group=1), data=ycs_cc) + 
  scale_y_date(limits=c(as.Date("1900-03-23"), as.Date("1900-06-04"))) +
  ylab("50th Quantile Arrival Date") +  facet_wrap(~LOC) 

ggplot(ycs_cc, aes(winterNDVI, sess.quant.50)) + geom_point() + 
  theme_bw() + stat_smooth(method="lm", formula = y ~ x, aes(group=1), data=ycs_cc) + 
  scale_y_date(limits=c(as.Date("1900-03-23"), as.Date("1900-06-04"))) +
  ylab("50th Quantile Arrival Date") +  facet_wrap(~LOC) 

#overall response
p3 <- ggplot(ycs_cc, aes(yearly.Te.10C.q, sess.quant.50)) + geom_point() + theme_classic() + theme(text = element_text(size=15)) + stat_smooth(method="lm", aes(group=1), col="black") + ylab("Timing of spring arrival\n") + xlab("")

p4 <- ggplot(ycs_cc, aes(winterNDVI, sess.quant.50)) + geom_point() + theme_classic() + theme(text = element_text(size=15)) + stat_smooth(method="lm", aes(group=1), col="black") + ylab("") + xlab("")

```

Load a function ('mod.evaluation') to evaluate hierarchical models
```{r}
source('/home/sarah/Documents/GitHub/extreme_limits/Model_evaluation_v4.r')
```

Using aforementioned function evaluate hierarchical models of median arrival as a linear function of annual physiological demand and resource availability, allowing for a random intercept associated with 'site' 
```{r}
attach(yearly.catch.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='sess.quant.50', centering='CGM', stand=T, ordered.fac.treatment = "as.num", log.D=T, R="winterNDVI") 
detach(yearly.catch.stat)
```

Using  function evaluate hierarchical models of 25th percentile arrival as a linear function of annual physiological demand and resource availability, allowing for a random intercept associated with 'site' 
```{r}
attach(yearly.catch.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='sess.quant.25', centering='CGM', stand=T, ordered.fac.treatment = "as.num", log.D=T, R="winterNDVI") 
detach(yearly.catch.stat)
```
Using  function evaluate hierarchical models of 75th percentile arrival as a linear function of annual physiological demand and resource availability, allowing for a random intercept associated with 'site' 
```{r}
attach(yearly.catch.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='sess.quant.75', centering='CGM', stand=T, ordered.fac.treatment = "as.num", log.D=T, R="winterNDVI") 
detach(yearly.catch.stat)
```