Model Molt Variation
========================================================

This document describes the model of Molt observed in Arizona and how it covaries with energetic demand and resource availability in Mexican wintering grounds

```{r}
library(ggplot2)
#load the Hummingbird data; i.e. the d.f. site.dat
load("/home/sarah/Documents/GitHub/extreme_limits/data/site.dat.rdata")
#keep only the variables you need
site.dat <- site.dat[, c("LOC", "PriMary.Molt", "yearfac", "mofac")]
#make an ordered categorical variable describing primary molt
ordermolt <- function(x){ordered(x, levels=c("M","R",1:8,0,9,"F","L"))}
site.dat$PriMary.Molt <- ordermolt(site.dat$PriMary.Molt)
hist(as.numeric(site.dat$PriMary.Molt),col="grey80", xlim = c(0,14), breaks = seq(0.5,14.5, by=1), main = "Primary Molt")
#plot molt stage by year for comparison
sd <- site.dat[complete.cases(site.dat),]
ggplot(sd, aes(PriMary.Molt)) + geom_histogram() + facet_wrap(~yearfac) + theme_bw() + xlab("Primary Molt Stage") + 
  ylab("number birds")
```



```{r}
#Calculate quantiles of molt distribution per site & year ('Molt.quant...'). quant.50 indicates the median molt stage at a given site and year

getMOLTquant <- function(quantile.){
  tapply(site.dat$PriMary.Molt, INDEX=list(site.dat$LOC, site.dat$yearfac),
        FUN=quantile, type=1, probs=c(quantile.), na.rm=T)}

MOLT.quant.05 <- ordermolt(levels(site.dat$PriMary.Molt)[as.vector(getMOLTquant(.05))])
MOLT.quant.25 <- ordermolt(levels(site.dat$PriMary.Molt)[as.vector(getMOLTquant(.25))])
MOLT.quant.50 <- ordermolt(levels(site.dat$PriMary.Molt)[as.vector(getMOLTquant(.50))])
MOLT.quant.75 <- ordermolt(levels(site.dat$PriMary.Molt)[as.vector(getMOLTquant(.75))])
MOLT.quant.95 <- ordermolt(levels(site.dat$PriMary.Molt)[as.vector(getMOLTquant(.95))])
MOLT.quants <- cbind.data.frame(MOLT.quant.05, MOLT.quant.25, MOLT.quant.50, MOLT.quant.75, MOLT.quant.95)
rm(MOLT.quant.05, MOLT.quant.25, MOLT.quant.50, MOLT.quant.75, MOLT.quant.95)

# calculate variables to hold the site names ('LOC') and years ('yr') associated with the quantiles
quant.temp <- tapply(site.dat$PriMary.Molt, INDEX=list(site.dat$LOC, site.dat$yearfac),
       FUN=quantile, type=1, probs=c(.05), na.rm=T)
LOC <- rownames(quant.temp)
yr <- rep(colnames(quant.temp), each=length(LOC))
rm(quant.temp)

# calculate the nr of birds in each sample ('n.birds') (number of birds at each site in each year)
n.birds <- as.vector(tapply(site.dat$PriMary.Molt, INDEX=list(site.dat$LOC, site.dat$yearfac),
                 FUN=function(x){length(which(!is.na(x)))}))

#save the new metrics in the d.f. "MOLT.quants"
MOLT.quants <- cbind.data.frame(LOC, yr, n.birds, MOLT.quants)
rm(n.birds, LOC, yr)
```

Plot the range, mean, and sd for each year for molt quant 50, for the manuscript
```{r}
mq50 <- MOLT.quants[,c(1:3,6)]
mq50 <- mq50[complete.cases(mq50),]
ggplot(mq50, aes(yr, MOLT.quant.50)) + geom_point(aes(size=n.birds)) + theme_bw()

moltscores <- table(mq50$MOLT.quant.50)
moltscores[11] / sum(moltscores)
moltscores[12] / sum(moltscores)
moltscores[13] / sum(moltscores)

# take a weighted average of the discrete classes for the mean molt score
num = sum(moltscores)
p = apply(moltscores, 1, function(x) {x/num} )
index = c(1:length(moltscores))

meanmolt = moltscores[round(sum(index * p))]
```

Load the yearly climate (and NDVI) data which will serve as predictors and merge it with the molt metrics in the data frame 'yearly.molt.stat' to be used for modelling
````{r}
load("/home/sarah/Documents/GitHub/extreme_limits/data/yearly.climate.rdata")
#gives you yearly.climate
yearly.molt.stat <- merge(x=MOLT.quants,y=yearly.climate,by='yr')
rm(MOLT.quants,yearly.climate)
````

XYPLOT VERSION: Plot molt responses to a proxy for physiological (yearly.Te.10C.q) demand and resrouce availability (NDVI)
```{r fig.width=7, fig.height=6, warning=FALSE}
require(lattice)
require(latticeExtra)
xyplot(MOLT.quant.50 ~ yearly.Te.10C.q|LOC, data=yearly.molt.stat) + layer(panel.smoother(x, y, method = "lm"))
xyplot(MOLT.quant.50 ~ winterNDVI|LOC, data=yearly.molt.stat) + layer(panel.smoother(x, y, method = "lm"))

#overall response
xyplot(MOLT.quant.50 ~ yearly.Te.10C.q, data=yearly.molt.stat) + layer(panel.smoother(x, y, method = "lm"))
xyplot(MOLT.quant.50 ~ winterNDVI, data=yearly.molt.stat) + layer(panel.smoother(x, y, method = "lm"))
```

GGPLOT VERSION: Plot molt responses to a proxy for physiological (yearly.Te.10C.q) demand and resrouce availability (NDVI)
```{r fig.width=7, fig.height=6, warning=FALSE}
require(ggplot2)

ggplot(yearly.molt.stat[complete.cases(yearly.molt.stat),], aes(yearly.Te.10C.q, MOLT.quant.50)) + geom_point() + theme_bw() + facet_wrap(~LOC) + stat_smooth(method="lm", aes(group=1))

ggplot(yearly.molt.stat[complete.cases(yearly.molt.stat),], aes(winterNDVI, MOLT.quant.50)) + geom_point() + 
  theme_bw() + facet_wrap(~LOC) + stat_smooth(method="lm", aes(group=1))

#overall response
p1 <- ggplot(yearly.molt.stat[complete.cases(yearly.molt.stat),], aes(yearly.Te.10C.q, MOLT.quant.50)) + geom_point() + theme_classic() + theme(text = element_text(size=15)) + stat_smooth(method="lm", aes(group=1), col="black") + ylab("Molt\n") + xlab("")

p2 <- ggplot(yearly.molt.stat[complete.cases(yearly.molt.stat),], aes(winterNDVI, MOLT.quant.50)) + geom_point() + theme_classic() + theme(text = element_text(size=15)) + stat_smooth(method="lm", aes(group=1), col="black") + ylab("") + xlab("")
```

Load a function ('mod.evaluation') to evaluate hierarchical models
````{r}
source('/home/sarah/Documents/GitHub/extreme_limits/Model_evaluation_v4.r')
````

Using  function evaluate hierarchical models of median Molt stage (50th Percentile of Molt) as a 
linear function of annual physiological demand and resource availability, allowing for a 
random intercept associated with 'site' 
Binomial, using winterNDVI as resource AVAILABILITY
````{r}
attach(yearly.molt.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='MOLT.quant.50', centering='CGM', stand=T, ordered.fac.treatment = "as.binomial", 
               log.D=T, R="winterNDVI", D="yearly.Te.10C.q")
detach(yearly.molt.stat)
````

Using function evaluate hierarchical models of 25th percentile Molt stage as a linear function of annual physiological demand and resource availability, allowing for a random intercept associated with 'site' 
Binomial, using winter NDVI as resource AVAILABILITY
````{r}
attach(yearly.molt.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='MOLT.quant.25', centering='CGM', stand=T, ordered.fac.treatment = "as.binomial", 
               log.D=T, R="winterNDVI", D="yearly.Te.10C.q")
detach(yearly.molt.stat)
````

Using  function evaluate hierarchical models of median Molt stage (75th Percentile of Molt) as a 
linear function of annual physiological demand and resource availability, allowing for a 
random intercept associated with 'site' 
Binomial, using winterNDVI as resource AVAILABILITY
````{r}
attach(yearly.molt.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='MOLT.quant.75', centering='CGM', stand=T, ordered.fac.treatment = "as.binomial", 
               log.D=T, R="winterNDVI", D="yearly.Te.10C.q")
detach(yearly.molt.stat)
````