Model Weight Variation
========================================================

This document describes the model of Weight observed in Arizona and how it covaries with energetic demand and resource availability in Mexican wintering grounds

```{r}
#load the Hummingbird data; i.e. the d.f. site.dat
load("/Users/sarah/Desktop/Dropbox/ActiveResearchProjects/Hummingbird_extreme_limits/site.dat.rdata")
```

How does weight vary with fat?
```{r}
#fat classes 0 and P can be lumped
site.dat$Fat[site.dat$Fat=="P"] <- 0
orderfat <- function(x){ordered(x, levels=c(0, "T" ,1:3))}
site.dat$Fat <- orderfat(site.dat$Fat)
plot(site.dat$Weight ~ site.dat$Fat, col="light grey", xlab = "Fat", ylab = "Weight")
```

How does weight vary with molt?
````{r}
ordermolt <- function(x){ordered(toupper(x), levels=c("M","R",1:8,0,9,"F","L"))}
site.dat$PriMary.Molt <- ordermolt(site.dat$PriMary.Molt)
plot(site.dat$Weight ~ site.dat$PriMary.Molt, col="light grey", xlab = "Primary Molt", ylab = "Weight")
````

FYI: How are observations distributed across Molt classes?
```{r}
table(site.dat$PriMary.Molt)
````


Which pairs of primary molt classes differ in weight?
````{r}
tt<-round(sort(TukeyHSD(aov(site.dat$Weight ~ site.dat$PriMary.Molt))[[1]][,4]), 3)
tt[tt<0.2]
````
Which pairs of secondary molt classes differ in weight?
````{r}
site.dat$Secondary.Molt <- ordermolt(site.dat$Secondary.Molt)
plot(site.dat$Weight ~ site.dat$Secondary.Molt, col="light grey")
table(site.dat$Secondary.Molt)
tt<-round(sort(TukeyHSD(aov(site.dat$Weight ~ site.dat$Secondary.Molt))[[1]][,4]), 3)
tt[tt<.2]
````
Which pairs of body molt classes differ in weight?
````{r}
site.dat$Body.Molt <- ordermolt(site.dat$Body.Molt)
plot(site.dat$Weight ~ site.dat$Body.Molt, col="light grey", xlab="Body Molt Class", ylab="Weight")
table(site.dat$Body.Molt)
tt<-round(sort(TukeyHSD(aov(site.dat$Weight ~ site.dat$Body.Molt))[[1]][,4]),3)
tt[tt<.2]
````
Which pairs of gorget/head molt classes differ in weight?
````{r}
site.dat$Gorget.head.molt <- ordermolt(site.dat$Gorget.head.molt)
plot(site.dat$Weight ~ site.dat$Gorget.head.molt, col="light grey",xlab="gorget head molt", ylab="Weight")
table(site.dat$Gorget.head.molt)
tt<-round(sort(TukeyHSD(aov(site.dat$Weight ~ site.dat$Gorget.head.molt))[[1]][,4]),3)
tt[tt<.2]
````
Which pairs of tail molt classes differ in weight?
````{r}
site.dat$Tail.Molt <- ordermolt(site.dat$Tail.Molt)
plot(site.dat$Weight ~ site.dat$Tail.Molt, col="light grey", xlab="tail molt", ylab="Weight")
table(site.dat$Tail.Molt)
tt<-round(sort(TukeyHSD(aov(site.dat$Weight ~ site.dat$Tail.Molt))[[1]][,4]),3)
tt[tt<.2]
```
```{r}
#keep only the variables you need for the modelling analysis
site.dat <- site.dat[,c("LOC", "Weight", "yearfac", "mofac")]
```

Load the yearly climate (and NDVI) data which will serve as predictors and merge it with the weight metrics in the data frame 'yearly.weight.stat' to be used for modelling
````{r}
load("/Users/sarah/Desktop/Dropbox/ActiveResearchProjects/Hummingbird_extreme_limits/yearly.climate.rdata")
#gives you yearly.climate
yearly.weight.stat <- merge(x=site.dat, y=yearly.climate, by.x='yearfac', by.y='yr')
rm(site.dat, yearly.climate)
````

Plot molt responses to a proxy for physiological (yearly.Te.10C.q) demand and resrouce availability (NDVI)
```{r fig.width=7, fig.height=6, warning=FALSE}
require(lattice)
require(latticeExtra)
xyplot(Weight ~ yearly.Te.10C.q|LOC, data=yearly.weight.stat) + layer(panel.smoother(x, y, method = "lm"))
xyplot(Weight ~ winterNDVI|LOC, data=yearly.weight.stat) + layer(panel.smoother(x, y, method = "lm"))
```

Load a function ('mod.evaluation') to evaluate hierarchical models
````{r}
source('/Users/sarah/Documents/GitHub/extreme_limits/Model_evaluation_v4.r')
````

Using aforementioned function evaluate hierarchical models of weight as a linear function of annual physiological demand and resource availability, allowing for a random intercept associated with 'site' 
````{r}
attach(yearly.weight.stat)
#http://www.quantpsy.org/interact/interactions.htm
mod.evaluation(yname='Weight', centering='CGM', stand=T, ordered.fac.treatment = "as.num", log.D=T, R="winterNDVI", D="yearly.Te.10C.q", N=NULL)

detach(yearly.weight.stat)
````
